Modelling the Lymphatic Metastatic Progression Pathways of OPSCC \\ from Multi-Institutional Datasets

### 2024-05-22 src/tex/ms.tex
[showyourwork] overleaf sync

We can use this assumption by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_\text{D} \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_\text{D} \mid \text{late} \right)$ for advanced T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions. They depend only on a single parameter, which when multiplied with the number of considered time-steps, conveniently also represents the distribution's mean.
%
\begin{equation} \label{eq:time_dist}
    P \left( t = t_\text{D} \mid \text{T}x \right) = \operatorname{\mathfrak{B}}(t_\text{max},p_{\text{T}x})
\end{equation}
%
Here, the parameter $p_{\text{T}x}$ can be interpreted as the probability that the patient with a tumor of T-category $x$ will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as the latest time-step $t_\text{max} = 10$, which will therefore give us a distribution over the diagnosis time that has its mean at $\mathbb{E}[t_\text{D}] = 10 \cdot p_{\text{T}x}$.

@@

{\it Contralateral Spread:} So far, our model concerns itself with the ipsilateral lymphatic spread. However, our data suggests that at least contralateral \gls{lnl} II is frequently involved in \gls{opscc} patients. It is also of interest clinically to be able to decide when the contralateral side may be spared entirely, as this could meaningfully reduce side effects such as xerostomia and dysphagia [citation needed]. Therefore extending the model to include some or all contralateral \glspl{lnl} would represent an important addition on its way to a clinical validation and ultimately implementation.

>>>>

We can use this assumption by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_\text{D} \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_\text{D} \mid \text{late} \right)$ for advanced T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions. They depend only on a single parameter, which when multiplied with the number of considered time-steps, conveniently also represents the distribution's mean.\footnote{In our previous publication on the lymphatic progression model, we present some experiments showing that the shape of the time-prior and number of time-steps have no impact as long as only one T-category is considered \cite{ludwig_hidden_2021}.}
%
\begin{equation} \label{eq:time_dist}
    P \left( t = t_\text{D} \mid \text{T}x \right) = \operatorname{\mathfrak{B}}(t_\text{max},p_{\text{T}x})
\end{equation}
%
Here, the parameter $p_{\text{T}x}$ can be interpreted as the probability that the patient with a tumor of T-category $x$ will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as the latest time-step $t_\text{max} = 10$, which will therefore give us a distribution over the diagnosis time that has its mean at $\mathbb{E}[t_\text{D}] = 10 \cdot p_{\text{T}x}$.

@@

{\it Contralateral Spread:} So far, our model concerns itself with the ipsilateral lymphatic spread. However, our data suggests that at least contralateral \gls{lnl} II is frequently involved in \gls{opscc} patients. It is also of interest clinically as it presents the opportunity for further reducing the \gls{ctv-n} when the predicted risk is low enough. Therefore extending the model to include the contralateral side could be an important next step on its way to a clinical validation and ultimately implementation.
